---
title: "Economic and Health Impact of Weather Events Across US"
output:
  html_document: default
---

# Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. For instance, many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data shows that tornadoes are the most dangerous weather event to the population health. The excessive heat is the second most harmful to population health event with high economic consequences. The economic impact of weather events was also analyzed. Flash floods and thunderstorm winds caused billions of dollars in property damages between 1950 and 2011. The largest crop damage caused by drought, followed by flood and hails.

# Data Processing
The analysis is performed on [Storm Events Database](https://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by [National Climatic Data Center](https://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). There is also some documentation of the data available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

To conduct the analysis the following steps are performed:

- Read and preprocess the data. Give a specific format to events with types Frost/Freeze, FROST/FREEZE and FROST\\FREEZE which refer to the same type of event.

```{r, echo = TRUE}
data <- read.csv('data/repdata%2Fdata%2FStormData.csv')
```

- look at the number of unique event types:
```{r, echo = TRUE}
length(unique(data$EVTYPE))
```

- change uppercase letters to lowercase:
```{r, echo = TRUE}
event_types <- tolower(data$EVTYPE)
```

- replace all punct. characters with a space and update the dataframe:
```{r, echo = TRUE}
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
data$EVTYPE <- event_types
```

This analysis focuses on two major events. First, it takes into account the dangerous events with respect to Population health. Second, the economic effects of weather events are analysed.

# Events related to Population Health

- the number of casualties are aggregated by the event type to find the events that are most harmful to population health
```{r, echo = TRUE}
library(plyr)
casualties <- ddply(data, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
```

- Find events that caused most death and injury
```{r, echo = TRUE}
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 5)

injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 5)
```

The top 5 events that caused largest number of deaths are:
```{r, echo = TRUE}
fatal_events[, c("EVTYPE", "fatalities")]
```

The top 5 events that caused most number of injuries are
```{r, echo = TRUE}
injury_events[, c("EVTYPE", "injuries")]
```

# Weather Events with considerable impact on the economy

The impact of weather events on the economy is anaysed by using the property damage and crop damage reportings/estimates.

- First,the property and crop damage for each event is analysed:
```{r, echo = TRUE}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```

```{r, echo = TRUE}
prop_dmg_exp <- sapply(data$PROPDMGEXP, FUN=exp_transform)
data$prop_dmg <- data$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(data$CROPDMGEXP, FUN=exp_transform)
data$crop_dmg <- data$CROPDMG * (10 ** crop_dmg_exp)
```

- Second, Compute the economic loss by event type:
```{r, echo = TRUE}
library(plyr)
econ_loss <- ddply(data, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))
```

- Third, filter out events that caused no economic loss:
```{r, echo = TRUE}
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 5)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 5)
```

The top 5 events that caused most property damage are the following:
```{r, echo = TRUE}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
```

The top 5 events that caused biggest crop damage are the following:
```{r, echo = TRUE}
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

# Results

1) Health impact of weather events

- The following plot shows the top deadly events in US in the period 1950-2011:
```{r, echo = TRUE}
library(ggplot2)

p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")

par(mfrow=c(2,1))
plot(p1)
plot(p2)
```
From the figures above, we can see that the tornadoes cause most number of deaths and injuries among all event types, followed by wind, flood and heat.

2) Economic impact of weather events

- The following plots show the weather cost to the US economy during the period 1950-2011:
```{r, echo = TRUE}
library(ggplot2)

p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")

par(mfrow=c(2,1))
plot(p1)
plot(p2)
```

From the above figures, we can see that flash floods and thunderstorm winds cost the largest property damages among weather-related natural diseasters. The most severe weather event in terms of crop damage is the drought.

Note that property damages are given in logarithmic scale due to large range of values. 
